---
title: "Explore data of Coux et al. 2016"
format:
  html:
    embed-resources: true
    code-fold: true
execute:
    message: false
    warning: false
---

```{r}
#| output: false
#| echo: false

library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
source(here::here("R", "analyse-coux.R"))
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(ggrepel)
library(bipartite)

guild_colors <- c(pollinator = "#2a78d6", plant = "#eb6834")
theme_set(theme_minimal(base_size = 11))
```

The data contains 5 different tables, that we can group in three categories:

1. plant-pollinator interaction list at each site
2. plant and pollinator abundances (one table per guild)
3. plant and pollinator traits (one table per guild)

We look at each of these categories in what follows.

## Interactions

Here is what the raw data looks like.

```{r}
data <- tar_read(data_coux)
head(data$interaction)
```

```{r}
#| echo: false
n_sites <- n_distinct(data$interaction$site)
n_pollinators <- n_distinct(data$interaction$pollinator)
n_plants <- n_distinct(data$interaction$plant)
prop_zero <- mean(data$interaction$links == 0)
```

The columns correspond to

- `site`: the community (`r n_sites` sites)
- `pollinator`: the pollinator species (`r n_pollinators` species across all sites)
- `plant`: the plant species (`r n_plants` species across all sites)
- `links`: the number of interaction events (visits) observed between that
  plant and that pollinator at that site (`r scales::percent(prop_zero)` show
zero observed visits).

::: {.callout-note}

`data_coux` is already cleaned by `clean_coux_data()`: species names are
lowercased everywhere (pollinator names are capitalised and plant names
inconsistently so in the raw data), and `alnus_glutinosa`/`alnus_serrulate` are
merged into a single `alnus_sp`. The raw interaction table has `alnus_glutinosa`
at site `blahamlak` with no matching trait/abundance row, and `alnus_serrulate`
at `crobirlee`/`crostelee` joining fine. Since no site uses both names, we treat
all three as the same plant recorded under two names. See `clean_coux_data()` in
`R/analyse-coux.R` for details; use `tar_read(data_coux_raw)` to see the
uncleaned version.

:::

We have coded a function to compute species level network metrics using the
`bipartite::specieslevel` function.

```{r}
net_metrics <- tar_read(net_metrics)
net_metrics
```

### Site richness

```{r}
site_richness <- data$interaction |>
  group_by(site) |>
  summarise(
    pollinator = n_distinct(pollinator),
    plant = n_distinct(plant),
    .groups = "drop"
  ) |>
  pivot_longer(
    c(pollinator, plant),
    names_to = "guild",
    values_to = "n_species"
  ) |>
  mutate(site = fct_reorder(site, n_species, .fun = sum))

ggplot(site_richness, aes(x = site, y = n_species, fill = guild)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = guild_colors) +
  labs(x = NULL, y = "Number of species", fill = NULL) +
  theme(panel.grid.major.y = element_blank(), legend.position = "top")
```

Sites range from a few species to much richer communities. Network size itself
is a potential confound to keep in mind in later analyses. We also note that
plant and pollinator richness are strongly positively correlated.

### Interaction matrices

```{r}
ggplot(data$interaction, aes(x = links)) +
  geom_histogram(
    binwidth = 1, fill = guild_colors[["pollinator"]],
    color = "white", linewidth = 0.2
  ) +
  scale_y_continuous(trans = "sqrt") +
  labs(
    x = "Observed visits (links)", y = "Number of pairs (sqrt scale)"
  )
```

### Degree distribution by guild

```{r}
ggplot(net_metrics, aes(x = guild, y = degree, fill = guild)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1) +
  scale_fill_manual(values = guild_colors) +
  labs(
    x = NULL, y = "Degree (number of partners)"
  ) +
  theme(legend.position = "none")
```

### An example network

The richest site, for a sense of what an actual interaction web looks like.

```{r}
#| fig-width: 8
#| fig-height: 6.5

web_list <- tar_read(web_list)
richest_site <- names(web_list)[which.max(vapply(web_list, sum, numeric(1)))]

plotweb(
  web_list[[richest_site]],
  higher_color = guild_colors[["pollinator"]],
  lower_color = guild_colors[["plant"]],
  link_color = "grey80",
  srt = 0,
  text_size = 1
)
```

## Abundances

We have plant presence/absence per site — despite the name, this is
binary (1/0), not a count.

```{r}
str(data$plant_abundance)
```

And pollinator abundances, which are actual counts.

```{r}
str(data$pollinator_abundance)
```

### Plant occurence

```{r}
n_sites_abundance <- n_distinct(data$plant_abundance$site)

plant_occupancy <- data$plant_abundance |>
  pivot_longer(-site, names_to = "plant", values_to = "present") |>
  group_by(plant) |>
  summarise(n_sites = sum(present), .groups = "drop") |>
  mutate(plant = fct_reorder(plant, n_sites))

ggplot(plant_occupancy, aes(x = plant, y = n_sites)) +
  geom_col(fill = guild_colors[["plant"]], width = 0.7) +
  coord_flip() +
  labs(
    x = NULL, y = paste0("Number of sites (out of ", n_sites_abundance, ")")
  )
```

Most plant species are recorded at the majority of sites, unlike
pollinators below, where a few generalists dominate and many species are
site-specific.

### Pollinator abundance

```{r}
pollinator_totals <- data$pollinator_abundance |>
  pivot_longer(-site, names_to = "pollinator", values_to = "abundance") |>
  group_by(pollinator) |>
  summarise(total_abundance = sum(abundance), .groups = "drop") |>
  mutate(pollinator = fct_reorder(pollinator, total_abundance))

ggplot(pollinator_totals, aes(x = pollinator, y = total_abundance)) +
  geom_col(fill = guild_colors[["pollinator"]], width = 0.7) +
  coord_flip() +
  scale_y_continuous(trans = "log1p", breaks = c(0, 10, 100, 1000, 10000)) +
  labs(
    x = NULL, y = "Total abundance across sites (log scale)"
  )
```

## Trait overview

We have plant traits: growth form, phenology, flower morphology and
nectar reward (14 plants).

```{r}
str(data$plant_trait)
```

We have pollinator traits: body size, foraging preferences, sociality
and larval diet (16 pollinators).

```{r}
str(data$pollinator_trait)
```

### Plant types and annual vs perennial

```{r}
ggplot(
  data$plant_trait,
  aes(x = fct_infreq(growth_form), fill = annual_perennial)
) +
  geom_bar() +
  labs(
    x = NULL, y = "Number of species", fill = NULL
  )
```

### Most pollinators are small-bodied and solitary

```{r}
ggplot(data$pollinator_trait, aes(x = body_length, fill = soc_sol)) +
  geom_histogram(binwidth = 2, color = "white", linewidth = 0.2) +
  labs(
    x = "Body length (mm)", y = "Number of species", fill = NULL
  )
```

## Trait spaces

We follow Coux et al. (2016)'s approach: traits are standardised and combined
into a single dissimilarity via Gower's distance, then ordinated with PCoA. To
see how each trait contributes to the two ordination axes, we fit every trait
separately onto the PCoA scores

### Plant trait space

```{r}
plant_trait_ready <- tar_read(plant_trait_ready)
plant_pcoa <- tar_read(plant_trait_pcoa)
plant_pct <- round(plant_pcoa$values$Rel_corr_eig[1:2] * 100, 1)

plant_pcoa_df <- as.data.frame(plant_pcoa$vectors[, 1:2])
names(plant_pcoa_df) <- c("Axis1", "Axis2")
plant_pcoa_df$plant <- rownames(plant_pcoa_df)
plant_pcoa_df$growth_form <- plant_trait_ready$growth_form

ggplot(plant_pcoa_df, aes(Axis1, Axis2, color = growth_form, label = plant)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PCoA1 (", plant_pct[1], "%)"),
    y = paste0("PCoA2 (", plant_pct[2], "%)"),
    color = NULL
  )
```

Trees cluster tightly together, while herbs are more spread out — reflecting
probably a more variable growth strategy within that group.

### Which traits drive the plant trait space

```{r}
plant_fit <- tar_read(plant_trait_fit)
plot_trait_importance(plant_fit)
```

```{r}
#| fig-width: 8
#| fig-height: 6.5

plot_trait_biplot(plant_pcoa, plant_fit, guild_colors[["plant"]])
```

Season splits the space along a diagonal (spring/winter-flowering species
vs. the rest), while growth form and inflorescence type separate species
mostly along PCoA1 — consistent with the tree/herb split seen above.

### Pollinator trait space

```{r}
pollinator_trait_ready <- tar_read(pollinator_trait_ready)
pollinator_pcoa <- tar_read(pollinator_trait_pcoa)
pollinator_pct <- round(pollinator_pcoa$values$Rel_corr_eig[1:2] * 100, 1)

pollinator_pcoa_df <- as.data.frame(pollinator_pcoa$vectors[, 1:2])
names(pollinator_pcoa_df) <- c("Axis1", "Axis2")
pollinator_pcoa_df$pollinator <- rownames(pollinator_pcoa_df)
pollinator_pcoa_df$soc_sol <- pollinator_trait_ready$soc_sol

ggplot(
  pollinator_pcoa_df,
  aes(Axis1, Axis2, color = soc_sol, label = pollinator)
) +
  geom_point(size = 2) +
  geom_text_repel(size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PCoA1 (", pollinator_pct[1], "%)"),
    y = paste0("PCoA2 (", pollinator_pct[2], "%)"),
    color = NULL
  )
```

### Which traits drive the pollinator trait space

```{r}
pollinator_fit <- tar_read(pollinator_trait_fit)
plot_trait_importance(pollinator_fit)
```

```{r}
#| fig-width: 8
#| fig-height: 6.5

plot_trait_biplot(pollinator_pcoa, pollinator_fit, guild_colors[["pollinator"]])
```

## Uniqueness and originality

### Uniqueness

We compute species uniqueness per species per site, using the `funrar` dedicated
function. Uniqueness is defined as

$$ \text{uniqueness}_i = \min_j d_{ij} $$

that is the minimal functional distance to another species present in the
community.

```{r}
plant_uniqueness <- tar_read(plant_uniqueness)
pollinator_uniqueness <- tar_read(pollinator_uniqueness)
all_uniqueness <- rbind(
  plant_uniqueness |> mutate(guild = "plant"),
  pollinator_uniqueness |> mutate(guild = "pollinator")
) |>
  mutate(species = fct_reorder(species, uniqueness, .fun = median))

ggplot(all_uniqueness, aes(x = species, y = uniqueness, fill = guild)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1) +
  scale_fill_manual(values = guild_colors) +
  facet_wrap(~guild, scales = "free_y") +
  coord_flip() +
  labs(x = NULL, y = "Functional uniqueness") +
  theme(legend.position = "none")
```

### Originality

Functional originality is the distance of a species from the trait centroid
of the community it belongs to, i.e. how much its traits differ from the
community's average:

$$ \text{originality}_i = \lVert x_i - \bar{x}_{\text{community}} \rVert $$

We compute it per species per site in the PCoA trait space, with the centroid
taken over the species present at each site.

```{r}
plant_originality <- tar_read(plant_originality)
pollinator_originality <- tar_read(pollinator_originality)
all_originality <- rbind(
  plant_originality |> mutate(guild = "plant"),
  pollinator_originality |> mutate(guild = "pollinator")
) |>
  mutate(species = fct_reorder(species, originality, .fun = median))

ggplot(all_originality, aes(x = species, y = originality, fill = guild)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1) +
  scale_fill_manual(values = guild_colors) +
  facet_wrap(~guild, scales = "free_y") +
  coord_flip() +
  labs(x = NULL, y = "Functional originality") +
  theme(legend.position = "none")
```

### Weighted originality

Above, the community centroid treats every present species equally. For
pollinators we also have abundance data, so we can instead weight the
centroid by how common each species is at a site.

```{r}
pollinator_originality_w <- tar_read(pollinator_originality_w)

originality_comparison <- rbind(
  pollinator_originality |> mutate(weighting = "unweighted"),
  pollinator_originality_w |> mutate(weighting = "weighted by abundance")
) |>
  mutate(
    weighting = factor(
      weighting,
      levels = c("unweighted", "weighted by abundance")
    ),
    species_within = reorder(
      paste(species, weighting, sep = "___"), originality,
      FUN = median
    )
  )

ggplot(
  originality_comparison,
  aes(x = species_within, y = originality, fill = weighting)
) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1) +
  scale_fill_manual(
    values = c(
      "unweighted" = "grey60",
      "weighted by abundance" = guild_colors[["pollinator"]]
    )
  ) +
  scale_x_discrete(labels = \(x) sub("___.*$", "", x)) +
  facet_wrap(~weighting, scales = "free_y") +
  coord_flip() +
  labs(x = NULL, y = "Functional originality") +
  theme(legend.position = "none")
```

We see that ranking shift with weighting. Common generalist, such `a_mellifera`
drop in originality as they pull the community weighted mean, while rarer
species such as `n_producta` increase in originality.

To check, to what extent species abundance drive their originality we plot
species weighted originality against their abundance for each site.

```{r}
pollinator_abundance <- data$pollinator_abundance |>
  pivot_longer(
    cols = a_mellifera:t_orthodoxus,
    names_to = "species",
    values_to = "abundance"
  ) |>
  filter_out(abundance == 0)

originality_abundance <- pollinator_originality_w |>
  left_join(pollinator_abundance, by = join_by(site, species))

ggplot(
  originality_abundance,
  aes(x = log10(abundance), y = originality, color = species),
) +
  geom_point() +
  labs(
    x = "Abundance (log10 scale)",
    y = "Weighted originality",
    color = "Species"
  )
```

We see that there is strong negative correlation between weighted originality
and species abundance. This means that we should be careful if we use this
metric.

### Originality vs uniqueness

Next, we want to check that species originality and uniqueness are not too
strongly correlated, to ensure that the two metrics are complementary
description of species position in the functional space.

```{r}
pol_originality_uniqueness <- pollinator_originality |>
  left_join(pollinator_uniqueness, by = join_by(site, species))

plant_originality_uniqueness <- plant_originality |>
  left_join(plant_uniqueness, by = join_by(site, species))

originality_uniqueness <- rbind(
  plant_originality_uniqueness |> mutate(guild = "plant"),
  pol_originality_uniqueness |> mutate(guild = "pollinator")
)

ggplot(
  originality_uniqueness, 
  aes(x = uniqueness, y = originality, color = guild)
) +
  geom_point() +
  labs(x = "Uniqueness", y = "Originality", color = "Guild")
```


