---
title: "Explore data of Coux et al. 2016"
format:
  html:
    embed-resources: true
    code-fold: true
execute:
    message: false
    warning: false
---

```{r}
#| output: false
#| echo: false

library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
source(here::here("R", "analyse-coux.R"))
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(ggrepel)
library(bipartite)

guild_colors <- c(pollinator = "#2a78d6", plant = "#eb6834")
theme_set(theme_minimal(base_size = 11))
```

The data contains 5 different tables, that we can group in three categories:

1. plant-pollinator interaction list at each site
2. plant and pollinator abundances (one table per guild)
3. plant and pollinator traits (one table per guild)

We look at each of these categories in what follows.

## Interactions

Here is what the raw data looks like.

```{r}
data <- tar_read(data_coux)
head(data$interaction)
```

```{r}
#| echo: false
n_sites <- n_distinct(data$interaction$site)
n_pollinators <- n_distinct(data$interaction$pollinator)
n_plants <- n_distinct(data$interaction$plant)
prop_zero <- mean(data$interaction$links == 0)
```

The columns correspond to

- `site`: the community (`r n_sites` sites)
- `pollinator`: the pollinator species (`r n_pollinators` species across all sites)
- `plant`: the plant species (`r n_plants` species across all sites)
- `links`: the number of interaction events (visits) observed between that
  plant and that pollinator at that site (`r scales::percent(prop_zero)` show
zero observed visits).

::: {.callout-note}

`data_coux` is already cleaned by `clean_coux_data()`: species names are
lowercased everywhere (pollinator names are capitalised and plant names
inconsistently so in the raw data), and `alnus_glutinosa`/`alnus_serrulate` are
merged into a single `alnus_sp`. The raw interaction table has `alnus_glutinosa`
at site `blahamlak` with no matching trait/abundance row, and `alnus_serrulate`
at `crobirlee`/`crostelee` joining fine. Since no site uses both names, we treat
all three as the same plant recorded under two names. See `clean_coux_data()` in
`R/analyse-coux.R` for details; use `tar_read(data_coux_raw)` to see the
uncleaned version.

:::

We have coded a function to compute species level network metrics using the
`bipartite::specieslevel` function.

```{r}
net_metrics <- tar_read(net_metrics)
net_metrics
```

### Sites vary widely in size

```{r}
site_richness <- data$interaction |>
  group_by(site) |>
  summarise(
    pollinator = n_distinct(pollinator),
    plant = n_distinct(plant),
    .groups = "drop"
  ) |>
  pivot_longer(
    c(pollinator, plant),
    names_to = "guild",
    values_to = "n_species"
  ) |>
  mutate(site = fct_reorder(site, n_species, .fun = sum))

ggplot(site_richness, aes(x = site, y = n_species, fill = guild)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = guild_colors) +
  labs(x = NULL, y = "Number of species", fill = NULL) +
  theme(panel.grid.major.y = element_blank(), legend.position = "top")
```

Sites range from a few species to much richer communities. Network size itself
is a potential confound to keep in mind in later analyses. We also note that
plant and pollinator richness are strongly positively correlated.

### Interaction matrices are sparse

```{r}
ggplot(data$interaction, aes(x = links)) +
  geom_histogram(
    binwidth = 1, fill = guild_colors[["pollinator"]],
    color = "white", linewidth = 0.2
  ) +
  scale_y_continuous(trans = "sqrt") +
  labs(
    x = "Observed visits (links)", y = "Number of pairs (sqrt scale)"
  )
```

### Degree distribution by guild

```{r}
ggplot(net_metrics, aes(x = guild, y = degree, fill = guild)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1) +
  scale_fill_manual(values = guild_colors) +
  labs(
    x = NULL, y = "Degree (number of partners)"
  ) +
  theme(legend.position = "none")
```

### An example network

The richest site, for a sense of what an actual interaction web looks like.

```{r}
#| fig-width: 8
#| fig-height: 6.5

web_list <- tar_read(web_list)
richest_site <- names(web_list)[which.max(vapply(web_list, sum, numeric(1)))]

plotweb(
  web_list[[richest_site]],
  higher_color = guild_colors[["pollinator"]],
  lower_color = guild_colors[["plant"]],
  link_color = "grey80",
  srt = 0,
  text_size = 1
)
```

## Abundances

We have plant presence/absence per site — despite the name, this is
binary (1/0), not a count.

```{r}
str(data$plant_abundance)
```

And pollinator abundances, which are actual counts.

```{r}
str(data$pollinator_abundance)
```

### Plant occurence is skewed

```{r}
n_sites_abundance <- n_distinct(data$plant_abundance$site)

plant_occupancy <- data$plant_abundance |>
  pivot_longer(-site, names_to = "plant", values_to = "present") |>
  group_by(plant) |>
  summarise(n_sites = sum(present), .groups = "drop") |>
  mutate(plant = fct_reorder(plant, n_sites))

ggplot(plant_occupancy, aes(x = plant, y = n_sites)) +
  geom_col(fill = guild_colors[["plant"]], width = 0.7) +
  coord_flip() +
  labs(
    x = NULL, y = paste0("Number of sites (out of ", n_sites_abundance, ")")
  )
```

Most plant species are recorded at the majority of sites, unlike
pollinators below, where a few generalists dominate and many species are
site-specific.

### Pollinator abundance is highly skewed

```{r}
pollinator_totals <- data$pollinator_abundance |>
  pivot_longer(-site, names_to = "pollinator", values_to = "abundance") |>
  group_by(pollinator) |>
  summarise(total_abundance = sum(abundance), .groups = "drop") |>
  mutate(pollinator = fct_reorder(pollinator, total_abundance))

ggplot(pollinator_totals, aes(x = pollinator, y = total_abundance)) +
  geom_col(fill = guild_colors[["pollinator"]], width = 0.7) +
  coord_flip() +
  scale_y_continuous(trans = "log1p", breaks = c(0, 10, 100, 1000, 10000)) +
  labs(
    x = NULL, y = "Total abundance across sites (log scale)"
  )
```

## Traits

We have plant traits: growth form, phenology, flower morphology and
nectar reward (14 plants).

```{r}
str(data$plant_trait)
```

We have pollinator traits: body size, foraging preferences, sociality
and larval diet (16 pollinators).

```{r}
str(data$pollinator_trait)
```

### Plant types and annual vs perennial

```{r}
ggplot(
  data$plant_trait,
  aes(x = fct_infreq(growth_form), fill = annual_perennial)
) +
  geom_bar() +
  labs(
    x = NULL, y = "Number of species", fill = NULL
  )
```

### Most pollinators are small-bodied and solitary

```{r}
ggplot(data$pollinator_trait, aes(x = body_length, fill = soc_sol)) +
  geom_histogram(binwidth = 2, color = "white", linewidth = 0.2) +
  labs(
    x = "Body length (mm)", y = "Number of species", fill = NULL
  )
```

## Species similarity from traits

We follow Coux et al. (2016)'s approach: traits are standardised and combined
into a single dissimilarity via Gower's distance, then ordinated with PCoA.
Binary traits that mark membership in one of several mutually-exclusive
categories (flowering season for plants, larval diet for pollinators) are
treated as *asymmetric* binary: two species that both lack a given season/diet
are not counted as similar for that reason. Ordinal scores (e.g. floral
fragrance, nectar amount) are treated as ordered factors rather than plain
numbers or nominal categories. See `prepare_plant_traits()`,
`prepare_pollinator_traits()` and `compute_trait_pcoa()` in `R/analyse-coux.R`
for the exact variable typing.

To see how each trait contributes to the two ordination axes, we fit every
trait separately onto the PCoA scores with `vegan::envfit()` (see
`fit_trait_ordination()` in `R/analyse-coux.R`) — this gives, per trait, an
$R^2$ (how well that trait's variation aligns with the ordination) and a
direction: a vector for numeric traits, or a centroid per level for
categorical ones. Fitting traits one at a time, rather than all together,
means a trait's missing values only cost that trait rows, not the whole
analysis. Plotting helpers (`plot_trait_importance()`, `plot_trait_biplot()`)
are also defined in `R/analyse-coux.R`.

### Plant trait space

```{r}
plant_trait_ready <- tar_read(plant_trait_ready)
plant_pcoa <- tar_read(plant_trait_pcoa)
plant_pct <- round(plant_pcoa$values$Rel_corr_eig[1:2] * 100, 1)

plant_pcoa_df <- as.data.frame(plant_pcoa$vectors[, 1:2])
names(plant_pcoa_df) <- c("Axis1", "Axis2")
plant_pcoa_df$plant <- rownames(plant_pcoa_df)
plant_pcoa_df$growth_form <- plant_trait_ready$growth_form

ggplot(plant_pcoa_df, aes(Axis1, Axis2, color = growth_form, label = plant)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PCoA1 (", plant_pct[1], "%)"),
    y = paste0("PCoA2 (", plant_pct[2], "%)"),
    color = NULL
  )
```

Trees cluster tightly together (large, single, actinomorphic
inflorescences), while herbs are more spread out — reflecting a genuinely
more variable growth strategy within that group.

### Which traits drive the plant trait space

```{r}
plant_fit <- tar_read(plant_trait_fit)
plot_trait_importance(plant_fit)
```

Flowering season (spring, winter), inflorescence type and growth form are
the strongest, significant contributors; several ordinal scores
(fragrance, nectar amount) sit just below the significance threshold.

```{r}
#| fig-width: 8
#| fig-height: 6.5

plot_trait_biplot(plant_pcoa, plant_fit, guild_colors[["plant"]])
```

Season splits the space along a diagonal (spring/winter-flowering species
vs. the rest), while growth form and inflorescence type separate species
mostly along PCoA1 — consistent with the tree/herb split seen above.

### Pollinator trait space

```{r}
pollinator_trait_ready <- tar_read(pollinator_trait_ready)
pollinator_pcoa <- tar_read(pollinator_trait_pcoa)
pollinator_pct <- round(pollinator_pcoa$values$Rel_corr_eig[1:2] * 100, 1)

pollinator_pcoa_df <- as.data.frame(pollinator_pcoa$vectors[, 1:2])
names(pollinator_pcoa_df) <- c("Axis1", "Axis2")
pollinator_pcoa_df$pollinator <- rownames(pollinator_pcoa_df)
pollinator_pcoa_df$soc_sol <- pollinator_trait_ready$soc_sol

ggplot(
  pollinator_pcoa_df,
  aes(Axis1, Axis2, color = soc_sol, label = pollinator)
) +
  geom_point(size = 2) +
  geom_text_repel(size = 3, show.legend = FALSE) +
  labs(
    x = paste0("PCoA1 (", pollinator_pct[1], "%)"),
    y = paste0("PCoA2 (", pollinator_pct[2], "%)"),
    color = NULL
  )
```

The two social species (`a_mellifera`, `b_terrestris`) sit apart from all
solitary species along PCoA1, which otherwise separates solitary species
mostly by body size and larval diet.

### Which traits drive the pollinator trait space

```{r}
pollinator_fit <- tar_read(pollinator_trait_fit)
plot_trait_importance(pollinator_fit)
```

Larval diet (`larv_plant`, `larv_necpol`), body weight, carrying structure
and sociality are the strongest, significant contributors — body length
and time of day matter much less.

```{r}
#| fig-width: 8
#| fig-height: 6.5

plot_trait_biplot(pollinator_pcoa, pollinator_fit, guild_colors[["pollinator"]])
```

The two social, corbiculate species (`a_mellifera`, `b_terrestris`) pull
strongly towards `larv_necpol` and away from the rest; `larv_plant` and
body weight define a second, largely independent direction among the
solitary species.

